 

### 2020-09-17 vertical-thermal-review.Rmd
incorporating Marielle's edits

**Authors (so far, not necessarily in final order):** 
Nidhi Vinod^1^,
Martijn Slot^2^,
Ian McGregor^3^,
Elsa Ordway,
Marielle Smith,
Tyeen Taylor,
Kristina J. Anderson-Teixeira^1,2^*

@@

@@

Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF). 
There is also increasing severity and frequency of heat waves (Perkins et al. 2012).
This is expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change (e.g., Bonan 2008).
With increases in temperature, temperate and tropical forest canopies are already functioning beyond their optimal photosynthetic threshold (Mau et al. 2020). 
This, alongside climate change-driven disturbances such as drought, may contribute to increasing tree mortality in forests around the world (McDowell et al. 2020; REFS).

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
Globally, we're seeing larger trees suffering more during drought (Benett et al. 2015; Stovall et al. 2019).
Increasing mortality of large trees, sometimes coupled with reduced tree growth rates, is driving a global trend towards more disturbed forests with younger trees (Nate McDowell et al.2020). 
Additionally, decrease in canopy cover or increase in canopy mortality as a result of increasing temperature (Cornelius Senf et al. 2018) has shown to increase understory temperatures which raises concern for understory species and forest biodiversity (Zellweger et al. 2019, Defrenne et al. 2013).

@@

Understanding thermal sensitivity in vertical tree gradient and height from top of the canopy to understory will provide crucial insights for how raising temperature affects forest ecology and biodiversity.

@@

*Our primary interest is the gradient in height and exposure from the top of the canopy to the understory in forests. However, because a lot of the relevant research has focused on exposure gradients near ground level (e.g., comparisons of sun and shade leaves), we also review studies focused on exposure gradients.*

>>>>

**Authors (so far, not necessarily in final order):** 
Nidhi Vinod^1^,
Martijn Slot^2^,
Ian McGregor^3^,
Elsa Ordway,
Marielle N. Smith^3^,
Tyeen Taylor,
Kristina J. Anderson-Teixeira^1,2^*

@@

4. Department of Forestry, Michigan State University, East Lansing, MI 48824, USA

@@

Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF). 
There is also increasing severity and frequency of heat waves (Perkins et al. 2012).
This is expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will have a critical influence on the future of trajectory of climate change (e.g., Bonan 2008).
With increases in temperature, temperate and tropical forest canopies are already or will be soon functioning beyond their optimal photosynthetic threshold (Mau et al. 2020, Huang et al. 2019). 
This, alongside climate change-driven disturbances such as drought, may contribute to increasing tree mortality in forests around the world (McDowell et al. 2020; Allen et al. 2015, Brienen et al. 2015).

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davis_microclimatic_2019].**
Globally, we're seeing larger trees suffering more during drought (Bennett et al. 2015; Stovall et al. 2019).
Increasing mortality of large trees, sometimes coupled with reduced tree growth rates, is driving a global trend towards more disturbed forests with younger trees (Nate McDowell et al.2020). 
Additionally, decrease in canopy cover or increase in canopy mortality as a result of increasing temperature (Cornelius Senf et al. 2018) has shown to increase understory temperatures which raises concern for understory species and forest biodiversity (Zellweger et al. 2019, Defrenne et al. 2013).

@@

Understanding thermal sensitivity along the vertical canopy profile, from the understory to the top of the canopy will provide crucial insights for how raising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions [with additional feedbacks to forest structure and microclimates].

@@

*Our primary interest is the gradient in height and exposure from the top of the canopy to the understory in forests. However, because a lot of the relevant research has focused on exposure gradients near ground level (e.g., comparisons of sun and shade leaves), and the implications [or utility] for ecosystem models we also review studies focused on exposure gradients.*
